---
title: "BRCANetworks"
date: "2026-04-10"

author: Nagashree Avabhrath, Mikhail Ukrainetz, Madison Moffett, Grant Smith, Lucia
  Williams, Mark Grimes
output:
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    theme: cosmo
    highlight: tango
  pdf_document:
    toc: true
    toc_depth: '4'
  word_document:
    toc: true
    toc_depth: '4'
vignette: >
  %\VignetteIndexEntry{PTMsToPathways}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, echo = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">>",
  fig.align = "center",
  fig.path = "plots/",
  fig.widthplots = 8,
  fig.height = 6,
  out.width = "100%",
  results = "hold"
)
```
Note that we will use the complete PPI and BioPlanet edgefiles available in "/Data_repository_path/" for this vignette.
Lucy todo: these can come with the package


This tutorial will load and create networks from the breast cancer cohort (BRCA, N = 122), produced by

> **Proteogenomic landscape of breast cancer tumorigenesis and targeted therapy**
   Krug, K., Jaehnig, E. J., Satpathy, S., Blumenberg, L., Karpova, A., Anurag, M., et al.
   *Cell* 183, 1436PTMsToPathways::name6.e31

And downloaded from Supplemental data S2 from

> **PhosphoDisco: A Toolkit for Co-regulated Phosphorylation Module Discovery in Phosphoproteomic Data**
   Schraink, T., Blumenberg, L., Hussey, G., George, S., Miller, B., Mathew, N., Gonzalez-Robles, T.J., Sviderskiy, V., Papagiannakopoulos, T., Possemato, R., et al.
   *Mol Cell Proteomics* 22, 100596. 10.1016/j.mcpro.2023.100596

### Preprocess data for PTMsToPathways functions
First, let's load the PTMsToPathways package so its functions are available:
```{r eval = TRUE}
library(PTMsToPathways)
```
The BRCA data table described above is provided with the PTMsToPathways package and and can be read in as follows:
```{r eval = TRUE}
file_path <- system.file("extdata", "PhosphoDiscoData_mmc9.txt", package = "PTMsToPathways")
newphos <- utils::read.table(file_path, header = TRUE,
                               stringsAsFactors = FALSE, sep = "\t", comment.char = "#",
                               na.strings = "", quote = "", fill = TRUE)
```
It has 4237 rows and 124 columns representing 4237 phosphosites and 122 samples
with phosphoproteomic data.
```{r eval = TRUE}
dim(newphos)
```
The first two columns are `gene_symbol` and `variable_sites_names` which we will
use to create row names for the PTM table to match the expected input format for
the PTMsToPathways functions. The [Raw Data Processing vignette](RawDataProcessing.html) gives another
example of how to process raw data tables to create the expected input format for the PTMsToPathways functions.
Here are the first few rows and columns of the data frame:
```{r eval = TRUE}
head(newphos[, 1:5])
```
Now we will process the first two columns to create row names for the PTM table.
We extract the amino acid and the site number from the `variable_sites_names` column
and remove trailing letters from the site number if they exist.
```{r eval = TRUE}
 newphos$Amino.Acid <- sapply(newphos$variable_sites_names, function(x) substring (x, 1, 1))
 newphos$Site <- trimws(substring(newphos$variable_sites_names, 2))
 newphos$Site <- sub("[a-z]$", "", newphos$Site)
 head(newphos[, c("gene_symbol", "variable_sites_names", "Amino.Acid", "Site")])
```
Now use the PTMsToPathways function `name.peptide` to create peptide names for
the row names of the PTM table.
```{r eval = TRUE} 
 newphos$Peptide.Name <- mapply(
   name.peptide, genes = newphos$gene_symbol,
   sites =  newphos$Site, aa = newphos$Amino.Acid)
```
Create `ptmtable` with PTMs as rows and samples as columns for use in the next
steps, and remove the columns we used to create the row names.
```{r eval = TRUE}
phosdata <- newphos[, 3:ncol(newphos), ]
rownames(phosdata) <- newphos$Peptide.Name
phosdata <- phosdata[, !(names(phosdata) %in% c("gene_symbol", "variable_sites_names", "Amino.Acid", "Site", "Peptide.Name"))]
ptmtable <- phosdata
head(ptmtable[, 1:5])
```
Next, we create clusters and networks from those clusters as in the [Creating Networks vignette](CreatingNetworks.html).

### Create Clusters and Co-Cluster Correlation Networks (CCCNs)
This takes about 10 minutes on a laptop, so we provide both the code and the pre-computed results for this step.
To re-run the analysis, run, the following:
```{r eval = FALSE}
set.seed(88)
clusterlist.data <- MakeClusterList(ptmtable,
                                    keeplength = 3, toolong = 3.5)
CCCN.data <- MakeCorrelationNetwork(adj.consensus.matrix,
                                    ptm.correlation.matrix)
```
Or load in pre-computed results from within the PTMsToPathways package:
```{r eval = TRUE}
clusterlist.data <- brca_clusterlist_data
CCCN.data <- brca_CCCN_data
```
Whether computed or loaded, the `cluster.data` and `CCCN.data` are lists that contain the following elements:
```{r eval = TRUE}
common.clusters <- clusterlist.data[[1]]
adj.consensus.matrix <- clusterlist.data[[2]]
ptm.correlation.matrix <- clusterlist.data[[3]]
ptm.cccn.edges <- CCCN.data[[1]]
gene.cccn.edges <- CCCN.data[[2]]
gene.cccn.nodes <- CCCN.data[[3]]
```
We expect 231 common clusters:
```{r eval = TRUE}
length(common.clusters)
```
If desired, the clusters can be trimmed to those > 4. This reduces the number of clusters
from 231 to 204.
```{r eval = TRUE}
cclength <- sapply(common.clusters, length)
common.clusters4 <- common.clusters[which(cclength>3)]
length(common.clusters4)
```
We can use [graph.ptm.by.cluster](references/graph.ptm.by.cluster.html) to visualize these in a heatmap. To demonstrate,
let's look at the output for the first 3 clusters. PTMs are rows and samples are columns, and color represents the value of the PTM
in that sample. Black indicates missing values.
```
```{r eval = TRUE}
fig_dir <- knitr::opts_current$get("fig.path")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
output <- file.path(fig_dir, "ptm_all_clusters_l4.pdf")
res <- graph.ptm.by.cluster(
     ptmtable         = ptmtable,
     common.clusters  = common.clusters4[1:3],          # use all clusters > 4, only first 3
     filename         = output,
     order.rows       = "slope",
     zlim             = 3,
     show.row.labels  = FALSE,
     show.col.labels  = TRUE,
     col_cex          = 0.7
   )
knitr::include_graphics(output)
```

PTMsToPathways provides the function [`EvaluateClusters`](references/EvaluateClusters.html)
which computes the following for each cluster:

* `intensity` = total signal after removing the NA fraction of samples for this cluster
* `realsamples` = samples that are not single-gene/PTM samples for this cluster
* `cleargenes` = genes/PTMs that fit a pattern that ranks by decreasing total signal
* `percent.NA` = percentage of missing values in the cluster sub-table

It also computes an `index` value for every cluster, which is a composite value computed from
the above. The output is ordered by this `index` value, so we examine the top 10 clusters here:
```{r eval = TRUE}
eval_brca <- EvaluateClusters(
   common.clusters4, ptmtable,
   data.type  = "ratio",
   use.slope  = FALSE,
   index.mode = "density",
   verbose    = FALSE
 )
eval_brca[1:10, ]

```
### Build Cluster Filtered Networks (CFNs) and Pathway Crosstalk Networks (PCNs)

For PPI edges, the code below demonstrates how to get the STRING-db and GeneMANIA edges from the static downloaded networks.
[todo: instructions to download once these are available.]
Alternatively, the resulting data can be loaded from within the package.

```{r eval = FALSE}
string_db_filepath <- "your/filepath/here.tsv"

stringdb.edges <- GetSTRINGdb.edges(gene.cccn.edges,
                              gene.cccn.nodes,
                              local                  = TRUE,
                              string.local.path      = string_db_filepath
)
```

To avoid downloading the STRINGdb edges, they can be loaded from within the package:
```{r eval = TRUE}
stringdb.edges <- BRCA_stringdb.edges
head(stringdb.edges)
```

Similary for GeneMANIA edges:
```{r eval = FALSE}  
genemania_db_filepath <- "your/filepath/here.tsv"
genemania.edges <- GetGeneMANIA.edges (gm.all.edges.path,
                                gene.cccn.nodes,
                                local                = TRUE,
                                genemania.local.path = genemania_db_filepath,
                                gm.interaction.types = c("Pathway", "Physical Interactions", "Predicted"))
```

And again, to avoid downloading the GeneMANIA edges, they can be loaded from within the package:
```{r eval = TRUE}
genemania.edges <- BRCA_genemania.edges
head(genemania.edges)
```

This file contains the following types of interactions:
```{r eval = TRUE}
unique(genemania.edges$interaction)
```

We choose all but "Genetic Interactions" to include in the following function. 
[todo: where is this filtering actually done?]

```{r eval = TRUE}
file_path <- system.file("extdata", "Kinase_Substrate_Dataset.txt", package = "PTMsToPathways")
kinsub.edges <- GetKinsub.edges(file_path, gene.cccn.nodes)

# Now we can build the CFN.
network.list <- BuildClusterFilteredNetwork(gene.cccn.edges,
                                            stringdb.edges,
                                            genemania.edges,
                                            kinsub.edges,
                                            db.filepaths = c())

combined.PPIs <- network.list[[1]]
cfn <- network.list[[2]]
dim(cfn)
# 67376 edges [todo, these are off, is it a problem?]
cfn.merged <- mergeEdges(cfn)
dim(cfn.merged)
# 27048 edges
```

We build the PCN from the BioPlanet pathways as done previously. This takes about a few minutes, so we provide both the code and the pre-computed results for this step.
To re-run the analysis, run the following:
```{r eval = FALSE}
bioplanet.file <- system.file("extdata", "bioplanet_pathway_June2025.csv", package = "PTMsToPathways")
PCN.data <- BuildPathwayCrosstalkNetwork(common.clusters, bioplanet.file,
                                         createfile = FALSE)
```
Or load in pre-computed results from within the PTMsToPathways package:
```{r eval = TRUE}
PCN.data <- BRCA_PCN.data
pathway.crosstalk.network <- PCN.data[[1]] # 679707 edges
PCNedgelist <- PCN.data[[2]]
pathways.list <- PCN.data[[3]]
dim(pathway.crosstalk.network)
```

Now we can explore these networks. 

### Preprocess modules from Shraink et al.

We will examine the modules from Schraink, et al., 2023, Supplemental Table S4

```{r eval = FALSE}
# Read in PhosphoDisco Modules file.
PD_module.file <- paste0(paperpath, "PhosDiscoModules_mmc11.txt")

PD_module.df <- utils::read.table(PD_module.file, header = TRUE,
                  stringsAsFactors = FALSE, sep = "\t", comment.char = "#",
                  na.strings = "", quote = "", fill = TRUE)
# Note these have trailing spaces as do the rownames(ptmtable)
# Column names: geneSymbol, variableSites, HDBSCAN.min_cluster_size.4
# 1017 rows
# Note differences from the PTM table imported above.
length(intersect(newphos$variable_sites_names, PD_module.df$variableSites)) # 530
length(intersect(newphos$gene_symbol, PD_module.df$geneSymbol)) # 161

# Make peptide names as above

PD_module.df$Amino.Acid <- sapply(PD_module.df$variableSites, function(x) substring (x, 1, 1))
PD_module.df$Site <- trimws(substring(PD_module.df$variableSites, 2))
PD_module.df$Site <- sub("[a-z]$", "", PD_module.df$Site)

PD_module.df$Peptide.Name <- mapply(
  name.peptide, genes = PD_module.df$geneSymbol,
  sites =  PD_module.df$Site, aa = PD_module.df$Amino.Acid)

# Treat modules like our clusters
PD_module.df2 <- PD_module.df[,c("Peptide.Name", "HDBSCAN.min_cluster_size.4")]
PD_module.list <- dlply(
  PD_module.df2,
  .(HDBSCAN.min_cluster_size.4),
  function(x) x$Peptide.Name
)
PD_module.genes.unique <- lapply(
  PD_module.list,
  function(x) unique(sub(" .*", "", x)))
```

### Compare P2P clusters and Schraink, et al. modules

P2P identified different clusters than these modules.
``` {r eval = FALSE}

mod63.interesct <- sapply(common.clusters, function(x) intersect(x, PD_module.list$`63`))

# Interesctions were found in common.clusters$ConsensusCluster1; common.clusters$ConsensusCluster3; common.clusters$ConsensusCluster4; common.clusters$ConsensusCluster5; common.clusters$ConsensusCluster6

# First, consider all ptms from major clusters that intersect with module 63
mod63.clust.ptms <- unlist(c(common.clusters$ConsensusCluster1, common.clusters$ConsensusCluster3, common.clusters$ConsensusCluster4, common.clusters$ConsensusCluster5, common.clusters$ConsensusCluster6))

# This code graphs the CFN/CCCN from all these PTMs:
funckey <- getFuncKey(paste0(Data_repository_path, "FunctionKey.txt"))
cfn.cccn <- ptms_to_cfn(mod63.clust.ptms, cfn = cfn.merged, pepsep = ";")
cfn_cccn.nodes <- make.cytoscape.node.file(cfn.cccn, funckey, ptmtable,
                                           include.gene.data = TRUE,
                                           include.coclustered.PTMs = TRUE)
g1 <- GraphCfn(cfn.edges = cfn.cccn, cfn.nodes = cfn_cccn.nodes,
               Network.title = "CFN/CCCN All PTMs 1", Network.collection = "PTMsToPathways")
 
```

This is a complex graph that shows PTMs clusters as cliques connected by yellow correlation edges surrounding a CFN of interconnected gene nodes. 

Let's focus on CDK1 substrates to complement the work done in Schraink, et al., 2023.

```{r eval = FALSE}
cdk1.kinsub <- filter.edges.1("CDK1", kinsub.edges) # 82 edges

cdk1.substrates <- cdk1.kinsub [which(cdk1.kinsub$source=="CDK1"), "target"]

cfn_cccn.nodes.cdksubs <- cfn_cccn.nodes[cfn_cccn.nodes$Gene.Name %in% cdk1.substrates, ]

```

Two ways to select for CDK1 substrates are presented.
```{r eval = FALSE}
# Method 1: Using RCy3.
selectNodes(cfn_cccn.nodes.cdksubs$id, by = "id", preserve=FALSE)
selectEdgesConnectingSelectedNodes()

createSubnetwork(nodes = getSelectedNodes(), edges = getSelectedEdges(), nodes.by.col = "id", edges.by.col = "name")

# Method 2: Using P2P functions.

cfn_cccn.nodes.cdksubs.edges <- filter.edges.0(nodenames = cfn_cccn.nodes.cdksubs$id, edge.file = cfn.cccn)

g2 <- GraphCfn(cfn.edges = cfn_cccn.nodes.cdksubs.edges, cfn.nodes = cfn_cccn.nodes.cdksubs,
               Network.title = "CFN/CCCN All PTMs 5", Network.collection = "PTMsToPathways")

# Gives the exact same network.
```

Now further simplify the graph to show only CCCN PTMs.
```{r eval = FALSE}

cfn_cccn.nodes.cdksubs <- cfn_cccn.nodes[cfn_cccn.nodes$Gene.Name %in% cdk1.substrates, ] # repeat of above to make this stand alone

cdksubs.genes <- unique(cfn_cccn.nodes.cdksubs$Gene.Name)
cdksubs.cfn <- filter.edges.0(cfn_cccn.nodes.cdksubs.genes, cfn.merged)
cdksubs.cfn.cccn <- get.co.clustered.ptms(cdksubs.cfn, ptm.cccn.edges=ptm.cccn.edges)
cdksubs.cfn.cccn.nodes <- make.cytoscape.node.file(cdksubs.cfn.cccn, funckey, ptmtable,
                                           include.gene.data = TRUE,
                                           include.coclustered.PTMs = TRUE)
g3 <- GraphCfn(cfn.edges = cdksubs.cfn.cccn, cfn.nodes = cdksubs.cfn.cccn.nodes,
               Network.title = "CFN/CCCN All PTMs 6", Network.collection = "PTMsToPathways")
```

Each of these graphs can be modified to set node size and shape using the following P2P functions that act on the front window in Cytoscape. Note the differences that reflect activation of different cell signaling pathways in tumors with different mutations: X02BR011 (AKT1 missense mutant); X21BR010 (PIK3CA missense mutant); and X05BR045 (TP53 nonsense and MLLT4 frameshift mutants). 

```{r eval = FALSE}
setNodeColorToRatios(plotcol="X03BR011")
setNodeColorToRowz(plotcol="X03BR011")  # This exaggerates the node size and shape somewhat. 
setNodeColorToRatios(plotcol="X21BR010")
setNodeColorToRowz(plotcol="X21BR010")    
setNodeColorToRatios(plotcol="X05BR045")
setNodeColorToRowz(plotcol="X05BR045")

# Note that different samples have dramatically different differences in PTMs that are up or down, which is reflected also in total in gene nodes.

```
