---
title: "Processing Raw Data for PTMsToPathways"
author: "Lucia Williams, Nagashree Avabhrath, Mikhail Ukrainetz, Madison Moffett, Grant Smith, Mark Grimes"
output:
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    theme: cosmo
    highlight: tango
  pdf_document:
    toc: true
    toc_depth: '4'
  word_document:
    toc: true
    toc_depth: '4'
vignette: >
  %\VignetteIndexEntry{Raw Data Processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, echo = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">>",
  fig.align = "center",
  fig.path = "plots/",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  results = "hold"
)
```

## Purpose

This vignette intends to help
users produce a matrix with PTM names as row names (e.g. `FYN p Y411`) and
numeric data in the columns.
The numeric values are the mass
spectrometer output, and `NA`s represent missing data rather than zeroes.
Ambiguous PTMs, where a PTM could match several proteins, must be separated by
semicolons (for example, `"AARS ubi k747; AMBLIL p U123"`).

Mass spectrometry data output will vary depending on the experimental
design, source of data, and software used to process the raw spectra. R
supports many file types and can automatically convert them into a data
frame. For example, `read.csv()` will take a csv file and convert it into
a data frame (`read.csv()` is a variation of `read.table()`). We start with
data in tab-delimited spreadsheet format. 

## Naming conventions

In this vignette, we use the following shorthand conventions when describing
PTMs based on the modifications present in the example data set. Your
data will dictate the names of modifications.

* *Gene.Name* = The HUGO Gene Name is used to identify the protein/gene
* *Phosphorylation* = "p"
* *Lysine acetylation* = "ack"
* *Lysine methylation* = "kme"
* *Arginine methylation* = "rme"
* *Ubiquitination* = "ubi"

## Preprocessing data

First, let's load the P2P package, since it contains some helpful pre-processing functions.
```{r eval = TRUE}
library(PTMsToPathways)
```

The example raw data file for this
vignette (downloadable [here](https://github.com/UM-Applied-Algorithms-Lab/PTMsToPathways/raw/refs/heads/main/inst/extdata/phospho_cleaned_mapped.txt),
or load directly into R using the commands below)
contains only phosphorylation sites.

To prepare data for input to PTMsTo pathways, we first read in the data file.
To use your own data file,
replace `file_path` variable with your own path to file, as in the commented line below.

```{r}
# file_path <- "path/to/your/file.txt"
file_path <- system.file("extdata", "phospho_cleaned_mapped.txt",
                         package = "PTMsToPathways")

newphos <- utils::read.table(file_path, sep = "\t", skip = 0, header = TRUE,
                             blank.lines.skip = T, fill = T, quote = "\"", dec = ".",
                             comment.char = "", stringsAsFactors = F)
dim(newphos)
```

As we can see, this table has 933 rows and 170 columns.

First remove internal control rows (reverse sequences), which should yield 908 remaining rows.

```{r}
newphos <- newphos[!is.na(newphos$AllGeneSymbols),]
dim(newphos)
```

Many investigators inspect data in Microsoft Excel, which can export
tab- or comma-delimited files. Unfortunately, Excel can silently convert
some gene names into dates when they appear in a cell by themselves. We
reverse that with the P2P helper function [`fix.excel()`](reference/fix.excel.html).
If there are dates in the `AllGeneSymbols` column, we can apply the `fix.excel()` function
to each of them to convert them back to gene names.
```{r}
newphos$AllGeneSymbols <- sapply(newphos$AllGeneSymbols, fix.excel)
```
Investigators will need to identify which columns contain key
information for analysis of PTMs. 

In this example, the key columns are:

- `Amino.Acid`: the modified amino acid, such as S or T
- `Positions.Within.Proteins`: the amino acid number in the protein sequence
- `Modification.Type`: the PTM class, such as phosphorylation
- `AllGeneSymbols`: the HUGO gene name(s) of the protein(s) containing the PTM, separated by ";" if more than one

```{r eval = TRUE}
newphos.header <- newphos[,c("AllGeneSymbols", "Amino.Acid",
                "Positions.Within.Proteins", "Modification.Type")]

head(newphos.header)
```
We provide another helper function, [`name.peptide()`](reference/name.peptide.html),
to handle ambiguous modification sites (a modification site
whose peptide sequence is the same in more than one protein) separated by ";" or
another separator.
```{r}
newphos.header$Peptide.Name <- mapply(
  name.peptide, genes = newphos$AllGeneSymbols,
  sites =  newphos$Positions.Within.Proteins, aa = newphos$Amino.Acid)
head(newphos.header)
```

### Data columns

Next we identify the data columns, which contain the string `Intensity`.
The example data file is from a multi-PTM study and the data in this
table are from just the phosphorylation pulldown (other tables are for
other PTM types). The optimal pulldown columns are straightforward to
identify by the pulldown strings present in the sample names, `pTyr` in
this case. They are also identifiable by zooming out and looking at the
patterns of missing data, the optimal pulldowns, as a group, have the
least missing data. In this data, the following abbreviations are used:
**C** = Crizotinib,
**D** = DMSO,
**E** = Erlotinib,
**Pr** = PR171.
So for example the first four columns can be interpreted as:
```{r eval = TRUE}
names(newphos)[grep("Intensity", names(newphos))][1]
```
*C1.1*: Crizotinib biological replicate 1- technical replicate 1
```{r eval = TRUE}
names(newphos)[grep("Intensity", names(newphos))][2]
```
*C1.2*: Crizotinib biological replicate 1- technical replicate 2
```{r eval = TRUE}
names(newphos)[grep("Intensity", names(newphos))][3]
```
*C2.1*: Crizotinib biological replicate 2- technical replicate 1
```{r eval = TRUE}
names(newphos)[grep("Intensity", names(newphos))][4]
```
*C2.2*: Crizotinib biological replicate 2- technical replicate 2

So, we first select columns that contain the string `Intensity`. There should be 108 of these:
```{r eval = TRUE}
data.cols <- newphos[,grep("Intensity", names(newphos))]
dim(data.cols)
```
And then select colums that have `pTyr`. There should be 36 of these:
```{r eval = TRUE}
data.cols <- newphos[,grep("pTyr", names(newphos))]
dim(data.cols)
```

Now simplify column names (remove `Intensity`):
```{r eval = TRUE}
names(data.cols) <- sapply(names(data.cols), function(x){
  unlist(strsplit(x, "Intensity."))[2]
  })
head(names(data.cols))
``` 

Make zero into `NA`, which it is. (Note that this may not apply if you are
confident that zero means actual zero, which is possible with certain technical
advances like DIA.) We should have 16265 `NA` values.
```{r eval = TRUE}
zer0 <- which(data.cols==0, arr.ind = TRUE)
data.cols <- replace (data.cols, zer0, NA)
sum(is.na(data.cols))
```

### Merge techincal replicates

These data have technical replicates, which means that the same samples
were run twice. Due to the stochastic selection of peptides for
detection, the pattern of missing values is slightly different between
technical replicates. We therefore merge the technical replicates taking
the value of either replicate where it's missing in the other, and
averaging values detected in both, using the P2P function [merge2cols()](reference/merge2cols.html).

Define technical replicates:
```{r eval = TRUE}
tr1.opt <- names(data.cols)[grep(".1", names(data.cols), fixed=TRUE)]
tr2.opt <- names(data.cols)[grep(".2", names(data.cols), fixed=TRUE)]
tr1.opt[1]
tr2.opt[1]
```

Use `merge2cols()` to average technical replicates. This function ignores NA
values in either column and takes the average in the case where there are two
values. In this case, the resulting data frame should have 18 total columns, one for each biological replicate.
```{r eval = TRUE}
phosdata.merged <- data.frame(matrix(nrow = nrow(data.cols), ncol = length(tr1.opt)))
for (i in seq_along(tr1.opt)) {
phosdata.merged[, i] <- merge2cols(
data.cols[, tr1.opt[i]],
data.cols[, tr2.opt[i]]
)}
dim(phosdata.merged)

```
And fix up the names of the merged columns to remove the technical replicate numbers:
```{r eval = TRUE}
names(phosdata.merged) <- sapply(tr1.opt, function(x){
  substr(x, start=1, stop=nchar(x)-2)
  })
names(phosdata.merged)[1]
```
Merge the header and the data together to make a complete data set:
```{r eval = TRUE}
phosdatafile <- cbind(newphos.header, phosdata.merged)
dim(phosdatafile)
```
This file could be saved for reference using `write.table()`:
```{r  eval = FALSE}
write.table(phosdatafile, file = "phosdatafile.txt",
            row.names = FALSE, sep = "\t")
```

For input into P2P, we want the row names to be the PTM names and the columns
just to be the merged technical replicate columns, so we just use that data and set the row names to
the `Peptide.Name` column:
```{r eval = TRUE}  
phosdata.df <- phosdata.merged
rownames(phosdata.df) <- phosdatafile$Peptide.Name
```
Notice that this data is the same as the `ex_small_ptm_table` that is provided with the package (though the column names are slightly
different):
```{r eval = TRUE}
all.equal(setNames(phosdata.df, NULL), setNames(ex_small_ptm_table, NULL))
```

Log base 2 transformation improves clustering.
```{r  eval = TRUE}
log2phosdata <- log2(phosdata.merged)
```

The [Creating Networks vignette](vignettes/GettingStarted.html) show how to use the functions provided in PTMsToPathways to analyze
data stored in a variable called `ptmtable`.
```{r eval = FALSE}
ptmtable <- log2phosdata
```

## Optional data processing steps

For experiments where treatment with drugs is compared to control
samples, adding treatment/control ratios as additional data column can
improve clustering. This optional step adds dimensions to the data set
that enhance focus on the changes in response to drug treatments.

 Simplify column names first:
```{r  Example ratio calculations}
names(phosdata.df) <- sapply(names(phosdata.df), function(x){
  paste(unlist(strsplit(x, "SEPTM_pTyr"))[1],
        unlist(strsplit(x, "SEPTM_pTyr"))[2], sep = "")
  })
```

Explore using ratios where control=rowMeans (D1, D2, D3):
```{r}
H3122control <- rowMeans(phosdata.df[, names(phosdata.df)
                                     [grep("H3122.D", names(phosdata.df))]],
                         na.rm=TRUE)
```

Change NaN to NA
```{r}
H3122control[is.nan(H3122control)] <- NA

PC9control <- rowMeans(phosdata.df[, names(phosdata.df)
                                   [grep("PC9.D", names(phosdata.df))]],
                       na.rm=TRUE)

PC9control[is.nan(PC9control)] <- NA
```

Calculate treatment/control ratios
```{r}
# H3122 cells
H3122.C1.ratio <- phosdata.df$H3122.C1/H3122control
H3122.C2.ratio <- phosdata.df$H3122.C2/H3122control
H3122.C3.ratio <- phosdata.df$H3122.C3/H3122control
H3122.PR1.ratio <- phosdata.df$H3122.PR1/H3122control
H3122.PR2.ratio <- phosdata.df$H3122.PR2/H3122control
H3122.PR3.ratio <- phosdata.df$H3122.PR3/H3122control

# PC9 cells
PC9.E1.ratio <- phosdata.df$PC9.E1/PC9control
PC9.E2.ratio <- phosdata.df$PC9.E2/PC9control
PC9.E3.ratio <- phosdata.df$PC9.E3/PC9control
PC9.PR1.ratio <- phosdata.df$PC9.PR1/PC9control
PC9.PR2.ratio <- phosdata.df$PC9.PR2/PC9control
PC9.PR3.ratio <- phosdata.df$PC9.PR3/PC9control
```

Put these columns in a data frame:
```{r}
phos_ratios <- data.frame(H3122.C1.ratio, H3122.C2.ratio, H3122.C3.ratio,
                          H3122.PR1.ratio, H3122.PR2.ratio, H3122.PR3.ratio,
                          PC9.E1.ratio, PC9.E2.ratio, PC9.E3.ratio,
                          PC9.PR1.ratio, PC9.PR2.ratio, PC9.PR3.ratio)
```
Check (should be `TRUE`):
```{r}  
identical(rownames(phosdata.df), rownames(phos_ratios))
```

Make limits to unweight extreme values. This has been shown to improve clustering, and a ratio of 1000 is biologically not really functional different than a ratio of 100. 
```{r}
hi.ratio <- which(phos_ratios >= 100, arr.ind = TRUE)
low.ratio <- which(phos_ratios <= 1/100, arr.ind = TRUE)
phos_ratios.lim <- replace (phos_ratios, hi.ratio, 100) 
phos_ratios.lim <- replace (phos_ratios.lim, low.ratio, 1/100) 
```
log2 transformation improves clustering:
```{r}
phos_ratios.lim.log2 <- log2(phos_ratios.lim)
phosdata_plus_ratios <- cbind(log2phosdata, phos_ratios.lim.log2)
```
And plot to check:

```{r}
boxplot(phosdata_plus_ratios)
```
And do one more check:
```{r}
identical(rownames(phos_ratios.lim.log2), rownames(log2phosdata)) 
```

This can be used as the example ptmtable for subsequent testing.
```{r}
# ptmtable <- phosdata_plus_ratios
```

## Combining data from multiple PTM experiments

For experiments involving multiple PTMs, or if investigators wish to
combine several data sets, the data can be combined. For example,
pulldowns were made to isolate acetylated and ubiquitinated peptides
from the same experimental samples. Combining the data is simply
repeating the above steps using the correct optimum data columns, then
making the column names the same, and binding all the rows together.

Suppose you had acetylation data in `ackdata.df` and ubiquitination data in
`ubidata.df`, both formatted as above for phosphorylation data in
`phosdata.df`. You could combine them as follows.

First, make sure the column names are the same:

```{r  eval = FALSE}
kgp <- phosdata.df
kga <- ackdata.df
kgu <- ubidata.df

names(kgp) <- sapply(names(kgp), function(x){
  paste(unlist(strsplit(x, "_pTyr"))[1], unlist(strsplit(x, "_pTyr"))[2],
        sep = "")
  })

names(kga) <- sapply(names(kga), function(x){
  paste(unlist(strsplit(x, "_AcK"))[1], unlist(strsplit(x, "_AcK"))[2],
        sep = "")
  })

names(kgu) <- sapply(names(kgu), function(x){
  paste(unlist(strsplit(x, "_Ubi"))[1], unlist(strsplit(x, "_Ubi"))[2],
        sep = "")
  })

identical(names(kgp), names(kga)) # Check TRUE
```

Then `rbind` them:
```{r  eval = FALSE}
ptmdata <- rbind (kgp, kga, kgu) 
```

Reorder:
```{r  eval = FALSE}
ptmdata <- ptmdata[order(rownames(ptmdata)),]
```

This optional step improves clustering in our hands:
```{r  eval = FALSE}  
log2ptmdata <- log2(ptmdata)
```

Finally, this dataframe could be used as the example `ptmtable` for the P2P functions.
```{r  eval = FALSE}
# ptmtable <- log2ptmdata
```
