---
title: "Analyzing Pathways from PTMs: A Guide"
author: "Nagashree Avabhrath, Mikhail Ukrainetz, Madison Moffett, Grant Smith, Lucia Williams, Mark Grimes"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: cosmo
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Mass_Spec_Data_Processing}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.path = "plots/",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  results = "hold"
)
```

<p> </p>

## Purpose

Mass spectrometry data output will vary depending on the experimental
design, source of data, and software used to process the raw spectra. We
start with data in tab-delimited spreadsheet format. (R has a variation
of read.table() called read.csv(), which can be used for comma-delimited
.csv files.) This vignette intends to help users produce a matrix with
PTM names as row names (e.g. FYN p Y411) and numeric data in all the
columns.

## Conventions

We use the following shorthand conventions when describing PTMs. This
can be modified if different modification types are under investigation.

*Gene.Name* = The HUGO Gene Name is used to identify the protein/gene

*Phosphorylation* = "p"

*Lysine acetylation* = "ack"

*Lysine methylation* = "kme"

*Arginine methylation* = "rme"

*Ubiquitination* = "ubi"

## Data input

Investigators will need to identify which columns contain key
information for analysis of PTMs. The example raw data file for this
vignette contains only phosphorylation sites (ex_phosphotable =
"phospho_cleaned_mapped.txt").

"Amino Acid" has the modified amino acid, e.g. S,T, etc.

"Positions Within Proteins" is the amino acid number in the protein
sequence.

Ambiguous modification sites (a modification site whose peptide sequence
is the same in more than one protein) have multiple possible postions
separated by ";".

"Modification Type" has Phosphorylation" etc.

Note that R converts spaces in column names to periods. After reading in
our data table, the relevant column names are:

**genes** = "AllGeneSymbols"

**positions** = "Positions.Within.Proteins"

**aa** = "Amino.Acid"

**modification** = "Modification.Type"

We provide a helper function *name.peptide()* to make a PTM name using
these columns. This function concatenates ambiguous modification sites
and ignores NA values that are present in cells or ambiguous gene names.

Many investigators will examine their data in Microsoft Excel, which can
export the spreadsheet in tab- or comma-delimited format. Unfortunately,
Excel inescapably turns some gene names into dates when present in a
cell by themselves. We reverse this with a helper function,
*fix.excel()*.

```{r read in first data sets for multiple PTMs}
# Helper functions
# Function to change dates back into gene names - Excel changes many genes into dates and this can't be turned off!
fix.excel <- function(cell) {
  fixgenes = c("CDC2", "1-Sep", "2-Sep", "3-Sep", "4-Sep", "5-Sep", "7-Sep", "8-Sep", "9-Sep", "10-Sep", "11-Sep", "15-Sep", "6-Sep", "1-Oct", "2-Oct", "3-Oct", "4-Oct", "6-Oct", "7-Oct", "11-Oct", "1-Mar", "2-Mar", "3-Mar", "4-Mar", "5-Mar", "6-Mar", "7-Mar", "8-Mar", "9-Mar", "10-Mar", "11-Mar", "C11orf58", 'C17orf57', 'C3orf10',  'C7orf51', "C11orf59", "C4orf16", "1-Dec", "14-Sep")
  corrects = c("CDK1", "SEPT1", "SEPT2", "SEPT3", "SEPT4", "SEPT5", "SEPT7", "SEPT8", "SEPT9", "SEPT10", "SEPT11", "SEPT15", "SEPT6", "POU2F1", "POU2F2", "POU5F1", "POU5F1", "POU3F1", "POU3F2", "POU2F3", "MARCH1", "MARCH2", "MARCH3", "MARCH4", "MARCH5", "MARCH6", "MARCH7", "MARCH8", "MARCH9", "MARCH10", "MARCH11", "SMAP", "EFCAB13", "BRK1", "NYAP1", "LAMTOR1", 'AP1AR', "DEC1", "SEPT14")
  cellv <- unlist(strsplit(as.character(cell), "; "))
  if (any(fixgenes %in% cellv)) {
    cellv.new <- gsub(fixgenes[fixgenes %in% cellv], corrects[fixgenes %in% cellv], cellv)
    return (paste(cellv.new, collapse="; "))
  } else return(cell)    }
# This function will handle ambiguous modification sites (a modification site whose peptide sequence is the same in more than one protein) separated by ";" or another separator
# Make peptide names using this function:
name.peptide <- function (genes, modification="p", sites, aa, pepsep=";")	{
  genes.v <- unlist(strsplit(genes, pepsep, fixed = TRUE))
  genes.v[which(genes.v == "NA")] <- NA
  genes.v <- genes.v[!is.na(genes.v)]
  sites.v <- unlist(strsplit(sites, pepsep, fixed = TRUE))
  sites.v[which(sites.v == "NA")] <- NA
  sites.v <- sites.v[!is.na(sites.v)]
  sites.v <- sapply(sites.v, function (x) paste (aa, x, sep=""))
  Peptide.v <- as.character(noquote(paste(genes.v[1:length(genes.v)], modification, sites.v[1:length(sites.v)], sep=" ")))
  Peptide <- paste(unique(Peptide.v), collapse="; ")
  return(Peptide)
}

# _________________________________________________________________________________
# Read in data file that is output from mass spectrometry. 

# You should replace file_path with your own path to file, as in the commented line below
file_path <- system.file("extdata", "phospho_cleaned_mapped.txt", package = "PTMsToPathways")
# file_path <- "path/to/your/file.txt"
newphos <- read.table(file_path, sep = "\t", skip = 0, header=TRUE, blank.lines.skip=T, fill=T, quote="\"", dec=".", comment.char = "", stringsAsFactors=F)

# First remove internal control rows (reverse sequences)
newphos <- newphos[!is.na(newphos$AllGeneSymbols),]

# If there are dates in the AllGeneSymbols column, use:
newphos$AllGeneSymbols <- sapply(newphos$AllGeneSymbols, fix.excel)

# Next, identify the columns that will be used to name the PTM

headercols <- c("AllGeneSymbols", "Amino.Acid", "Positions.Within.Proteins", "Modification.Type")

# Make a separate data frame using these columns. 
newphos.head <- newphos[,headercols]

# Now create a unique PTM name (peptide name)
newphos.head$Peptide.Name <- mapply(name.peptide, genes=newphos.head$AllGeneSymbols, sites= newphos.head$Positions.Within.Proteins, aa=newphos.head$Amino.Acid)
```

Next we identify the data columns, which contain the string "Intensity."
The example data file is from a multi-PTM study and the data in this
table are from just the phosphorylation pulldown (other tables are for
other PTM types). The optimal pulldown columns are straightforward to
identify by the pulldown strings present in the sample names, "pTyr" in
this case. They are also identifiable by zooming out and looking at the
patterns of missing data, the optimal pulldowns, as a group, have the
least missing data. The second step is to select colums that have
"pTyr."

These data have technical replicates, which means that the same samples
were run twice. Due to the stochastic selection of peptides for
detection, the pattern of missing values is slightly different between
technical replicates. We therefore merge the technical replicates taking
the value of either replicate where it's missing in the other, and
averaging values detected in both, using the function, *merge2cols()*.

## Key for column names

**C** = Crizotinib

**D** = DMSO

**E** = Erlotinib

**Pr** = PR171

***For Intensity columns***

*C1-1*: Crizotinib biological replicate 1- technical replicate 1

*C1-2*: Crizotinib biological replicate 1- technical replicate 2

*C2-1*: Crizotinib biological replicate 2- technical replicate 1

*C2-2*: Crizotinib biological replicate 2- technical replicate 2

*C3-1*: Crizotinib biological replicate 3- technical replicate 1

*C3-2*: Crizotinib biological replicate 3- technical replicate 2

... & similar for other drugs

```{r reformat data sets for multiple PTMs}
# Define Data columns
phosdata <- newphos[,grep("Intensity", names(newphos))]
phosdata <- phosdata[,grep("pTyr", names(phosdata))]

# Simplify column names
names(phosdata) <- sapply(names(phosdata), function (x) unlist(strsplit(x, "Intensity."))[2])

# Make zero into NA, which it is. (Note that this may not apply if you are confident that zero means actual zero, which is possible with certain technical advances.) 
zer0 <- which(phosdata==0, arr.ind = TRUE)
phosdata <- replace (phosdata, zer0, NA)

# Define technical replicates
tr1.opt <- names(phosdata)[grep(".1", names(phosdata), fixed=TRUE)]
tr2.opt <- names(phosdata)[grep(".2", names(phosdata), fixed=TRUE)]

# Use this function to average technical replicates. This function ignores NA values in either column and takes the average in the case where there are two values.
merge2cols <- function (colv1, colv2) {
  newcolv=NA
  if (is.na(colv1) & is.na(colv2)) {
    newcolv=NA
    return(newcolv)} else
      if (is.na(colv1) | is.na(colv2)) {
        newcolv <- sum(colv1, colv2, na.rm=TRUE)
        return(newcolv) } else
          newcolv <- (colv1 + colv2)/2
        return(newcolv) }

# Use merge2cols() function to average technical replciates:
phosdata.merged <- data.frame(matrix(nrow=nrow(phosdata), ncol=18))
for(i in 1:length(tr1.opt)) {
  phosdata.merged[,i] <- mapply(merge2cols, colv1=as.numeric(phosdata[, tr1.opt[i]]), colv2=as.numeric(phosdata[,tr2.opt[i]]))
}
names(phosdata.merged) <- sapply(tr1.opt, function(x) substr(x, start=1, stop=nchar(x)-2))

# Merge with header
phosdatafile <- cbind(newphos.head, phosdata.merged)
# This file could be safed for reference using write.table()
write.table(phosdatafile, file="phosdatafile.txt", row.names = FALSE, sep="\t")

# For subsequent steps
rownames(phosdatafile) <- phosdatafile$Peptide.Name
phosdata.df <- phosdatafile[,6:23]
# This contains just the data with individual PTMs as rownames

# Log base 2 transformation improves clustering.
log2phosdata <- log2(phosdata.df)

# This could be used as the ptmtable for subsequent clustering.
# ptmtable <- log2phosdata
```

For experiments where treatment with drugs is compared to control
samples, adding treatment/control ratios as additional data column can
improve clustering. This optional step adds dimensions to the data set
that enhance focus on the changes in response to drug treatments.

```{r  Example ratio calculations}

# Simplify column names first:
names(phosdata.df) <- sapply(names(phosdata.df), function (x) paste(unlist(strsplit(x, "SEPTM_pTyr"))[1], unlist(strsplit(x, "SEPTM_pTyr"))[2], sep=""))

# Explore using ratios where control=rowMeans (D1, D2, D3)

H3122control <- rowMeans(phosdata.df[, names(phosdata.df)[grep("H3122.D", names(phosdata.df))]], na.rm=TRUE)
# Change NaN to NA
H3122control[is.nan(H3122control)] <- NA

PC9control <- rowMeans(phosdata.df[, names(phosdata.df)[grep("PC9.D", names(phosdata.df))]], na.rm=TRUE)
# Change NaN to NA
PC9control[is.nan(PC9control)] <- NA

# Calculate treatment/control ratios
# H3122 cells
H3122.C1.ratio <- phosdata.df$H3122.C1/H3122control
H3122.C2.ratio <- phosdata.df$H3122.C2/H3122control
H3122.C3.ratio <- phosdata.df$H3122.C3/H3122control
H3122.PR1.ratio <- phosdata.df$H3122.PR1/H3122control
H3122.PR2.ratio <- phosdata.df$H3122.PR2/H3122control
H3122.PR3.ratio <- phosdata.df$H3122.PR3/H3122control
# PC9 cells
PC9.E1.ratio <- phosdata.df$PC9.E1/PC9control
PC9.E2.ratio <- phosdata.df$PC9.E2/PC9control
PC9.E3.ratio <- phosdata.df$PC9.E3/PC9control
PC9.PR1.ratio <- phosdata.df$PC9.PR1/PC9control
PC9.PR2.ratio <- phosdata.df$PC9.PR2/PC9control
PC9.PR3.ratio <- phosdata.df$PC9.PR3/PC9control

# Put these columns in a data frame
phos_ratios <- data.frame(H3122.C1.ratio, H3122.C2.ratio, H3122.C3.ratio, H3122.PR1.ratio, H3122.PR2.ratio, H3122.PR3.ratio, PC9.E1.ratio, PC9.E2.ratio, PC9.E3.ratio, PC9.PR1.ratio, PC9.PR2.ratio, PC9.PR3.ratio)
# Check
identical(rownames(phosdata.df), rownames(phos_ratios))
# TRUE
# Make limits to unweight extreme values. This has been shown to improve clustering, and a ratio of 1000 is biologically not really functional different than a ratio of 100. 
hi.ratio <- which(phos_ratios>=100, arr.ind = TRUE)
low.ratio <- which(phos_ratios <=1/100, arr.ind = TRUE)
phos_ratios.lim <- replace (phos_ratios, hi.ratio, 100) 
phos_ratios.lim <- replace (phos_ratios.lim, low.ratio, 1/100) 
# log2 transformation improves clustering 
phos_ratios.lim.log2 <- log2(phos_ratios.lim)
# lucy comment: this was below but we needed phosdata_plus_ratios
phosdata_plus_ratios <- cbind(log2phosdata, phos_ratios.lim.log2)

boxplot(phosdata_plus_ratios)
# Now merge intensity and ratio data
# Check
identical(rownames(phos_ratios.lim.log2), rownames(log2phosdata)) 
# TRUE, okay


# This can be used as the example ptmtable for subsequent testing.
# ptmtable <- phosdata_plus_ratios
```

For experiments involving multiple PTMs, or if investigators wish to
combine several data sets, the data can be combined. For example,
pulldowns were made to isolate acetylated and ubiquitinated peptides
from the same experimental samples. Combining the data is simply
repeating the above steps using the correct optimum data columns, then
making the column names the same, and binding all the rows together.

```{r  Subsequent steps to merge datasets from a mulit-PTM experiment}
######################
# Workflow: Repeat the with aceltylation or other PTM data (ubiquitination in this case).
# This can be done with data that include ratios as above if desired.
# How to put them together:
# Make column names common. For example:
kgp <- phosdata.df
# lucy comment: ackdata.df not defined, so I'm just commenting everything out below
# kga <- ackdata.df
# kgu <- ubidata.df
# names(kgp) <- sapply(names(kgp), function (x) paste(unlist(strsplit(x, "_pTyr"))[1], unlist(strsplit(x, "_pTyr"))[2], sep=""))
# names(kga) <- sapply(names(kga), function (x) paste(unlist(strsplit(x, "_AcK"))[1], unlist(strsplit(x, "_AcK"))[2], sep=""))
# names(kgu) <- sapply(names(kgu), function (x) paste(unlist(strsplit(x, "_Ubi"))[1], unlist(strsplit(x, "_Ubi"))[2], sep=""))
# identical(names(kgp), names(kga)) # Check TRUE
#
# ptmdata <- rbind (kgp, kga, kgu) 
# Reorder here
# ptmdata <- ptmdata[order(rownames(ptmdata)),]

# This optional step improves clustering in our hands:
# log2ptmdata <- log2(ptmdata)

# This could be used as the ptmtable for clustering.
# ptmtable = log2ptmdata
```
